---
title: "Mastering the Process of Data Science - Assignment"
author: "Patrik Szigeti"
date: '06/09/2019'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Preparation

The following packages were used to perform data wrangling: `dplyr`, `tibble`, `stringr` and `readxl`.

- **Drinks.csv**

  - A new variable called `total_liters_of_pure_alcohol` was created based on alcohol consumption (`?`s treated as `NA`s) and the corresponding ounces and alcohol levels. Ounces were converted to liters.
  - Rows with `NA`s were omitted.

- **LifeExpectancy.csv**

  - I omitted the records without information regarding `income_group` and transformed the values to a factor with readable levels (removed the `_income` suffix).
  - I added 60 to the `Life expectancy at age 60 (years)` metric to have all three life expectancy metrics on the same level.
  - Out of the three metrics regarding expected lifetime, I created a fourth one, called `avg_life_expectancy` that takes the mean of the values.

- **CountriesOfTheWorld.xls**

  - Headers spreaded over two rows in the spreadsheet, and the table didn't start in the first row, but the fourth.
  - As some of the variables (`literacy`, `arable`, `crops`, `other`) were percentages, these were divided by 100.
  
## Analysis

```{r, include = FALSE}
# Libraries
library(corrr)
library(ggplot2)

# Load the clean datasets
source("data-preparation.R")

# Merge the datasets
data <- drinks %>% 
  inner_join(lifetime, by = "country") %>% 
  inner_join(countries, by = "country") %>% 
  mutate(country = as.factor(country)) %>% 
  arrange(country, year)

# Remove individual datasets
rm(drinks, lifetime, countries)
```

The following packages were used to perform analysis: `corrr` and `ggplot2`.

I joined the datasets on `country`.

### Correlation between expected lifetime and total liters of pure alcohol

```{r, message = FALSE}
res <- data %>% 
  select_if(is.numeric) %>%  
  correlate() %>% 
  focus(avg_life_expectancy)

res %>% 
  filter(rowname == "total_liters_of_pure_alcohol")
```

The correlation between life expectancy and total liters of pure alcohol consumed is relatively high (`0.473`). This suggests that the more alcohol someone consumes, the longer they might live. While the positive correlation is clear, I wouldn't necessarily agree with the hypothesis, as this would mean that drinking more can lead to longer lifetime. But are there any other factors that are highly correlated with life expectancy?

### Factors that are highly correlated to the expected lifetime

```{r, fig.align = "center"}
res %>% 
  mutate(rowname = reorder(rowname, avg_life_expectancy),
         sign = ifelse(avg_life_expectancy >= 0, "positive", "negative")) %>% 
  ggplot(aes(x = rowname, y = avg_life_expectancy, fill = sign)) +
    geom_bar(stat = "identity") +
    scale_fill_manual(values = c("positive" = "palegreen3", "negative" = "indianred2")) +
    labs(y = "Correlation with expected lifetime", x = "Factor") +
    coord_flip() +
    theme(legend.position = "none")
```

We can draw some conclusions based on this chart:

- All three of the alcohol servings seem to have a large, positive impact on the expected lifetime. This makes sense considering these three make up the value for `total_liters_of_pure_alcohol`, but the causation is questionable here as well. Although some of it would coincide with traditional home remedies, a glass of wine is good for circulation, while a shot of *pálinka* (Hungarian spirit) is said to cure all illnesses.
- `phones` has the highest positive correlation with life expectancy, meaning if there are more phones in a country, people will live longer. This makes sense since citizens of wealthier countries typically have more phones than those in poor ones.
- `literacy` and `gdp_per_capita` are usually also associated with wealthier countries where due to the better circumstances, people tend to live longer.
- The three highest negatively correlated factors also make a lot of sense:
  - `deathrate` or mortality rate is the scaled number of deaths in a particular population, which means when more people die, the life expectancy is lower.
  - `birthrate` is the total number of live births per 1000 in a year, and if this number is smaller, there are more deaths, which results in lower expected lifetime.
  - `infant_mortality_rate` is the death of young children under the age of 1, where the same principle applies as it did for `birthrate`.
- One interesting thing to see is the `year` variable's negative correlation with life expectancy, which is alarming. This means that life expectancy was higher in earlier readings of the metric, while lately it has been in decline.

## Summary

Overall we can say that while it seems like alcohol consumption has a relatively high positive effect on life expectancy, common sense dictates that this is not the case. Variables where higher values are usually associated with wealthier countries (such as `phones`, `literacy` and `gdp_per_capita`) are seemingly more important and are positively correlated with life expectancy. The same can be said for the opposite direction, higher `deathrate`, `birthrate` and `infant_mortality_rate` are typically connected to poor countries. However, deciding whether this assumption is correct is not part of this analysis.

**Suggestions:**

- Analyze whether regions and income groups have the assumed effect on life expectancy.
- Look into why expected lifetime seems to be in decline as years go by, and whether it's due to the same reasons as for the point above.

___

*This documentation and the codes are also available in my [ceu-mtpods GitHub repository](https://github.com/szigony/ceu-mtpods).*
